An OpenCL-based tridiagonal solver for evaluation of compact finite differences on parallel architectures

### 2015-11-24 thesis/acknowledgments.tex
some more updates to thesis

I owe my deepest thanks
to my advisor, Dr. Richard S. Miller,
for hist immense guidance and support;
and the members of my committee,
Dr. Gang Li and Dr. Lonny Thompson
for their valuable time.
I would like to acknowledge
Dr. Daniel Livescu at Los Alamos National Lab
for introducing us to this interesting problem,
for his helpful comments,
and for generous funding.
I also thank
Dr. Melissa Smith and Karan Sapra
for their helpful guidance,
and for organizing an excellent course on GPU computing.
I owe a great deal of gratitude to
Dr. Christopher Cox at Clemson University
for his guidance on a range of topics
and for many interesting conversations.
I would like to thank the
staff of the Cyberinfrastructure Technology Integration,
and especially
Dr. Galen Collier and Dr. Marcin Ziolkowski
for their valuable comments and support.
I owe thanks to Clemson University for the generous
allocation of compute time on the Palmetto cluster.
I acknowledge
Dr. K. N. Seetharamu and Dr. V. Krishna,
my advisors at PES Institute of Technology
for inspiring in me a passion for
scientific research.
I am indebted to Dr. Anush Krishnan
for introducing me to the fascinating field of
scientific computing.
I owe a special thanks to the
Software Carpentry Foundation
and its members,
for helping me be
a more productive programmer,
a more deliberate learner,
a more effective teacher,
and a better person.

Finally,
I would like to thank my family and friends
for everything they have done for me.

>>>>

I owe my deepest thanks
to my advisor, Dr. Richard S. Miller,
for his immense guidance and support;
and the members of my committee,
Dr. Gang Li and Dr. Lonny Thompson
for their valuable time.
I would like to acknowledge
Dr. Daniel Livescu at Los Alamos National Lab
for introducing us to this interesting problem,
for his helpful comments,
and for generous funding.
I also thank
Dr. Melissa Smith and Karan Sapra
for their helpful guidance,
and for organizing an excellent course on GPU computing.
I owe a great deal of gratitude to
Dr. Christopher Cox at Clemson University
for his guidance on a range of topics
and for many interesting conversations.
I would like to thank the
staff of the Cyberinfrastructure Technology Integration group at Clemson,
and especially
Dr. Galen Collier and Dr. Marcin Ziolkowski
for their valuable comments and support.
I owe thanks to Clemson University for the generous
allocation of compute time on the Palmetto cluster.
I acknowledge
Dr. K. N. Seetharamu and Dr. V. Krishna,
my advisors at PES Institute of Technology
for inspiring in me a passion for
scientific research.
I am indebted to Dr. Anush Krishnan
for introducing me to the fascinating field of
scientific computing.
I owe a special thanks to the
Software Carpentry Foundation
and its members,
for helping me be
a more productive programmer,
a more deliberate learner,
a more effective teacher,
and a better person.

Finally, I thank my family and friends
for everything they have done for me.

### 2015-11-24 thesis/introduction-and-background.tex
some more updates to thesis

When solving for flows numerically,
at each of the grid points in the computational domain,
various data about the flow must be stored---for
example, the geometric coordinates (x, y and z),
pressure, temperature and velocities at that point.
For even small problem sizes,
the amount of memory required to store this data
can quickly exceed the capacity of modern workstations/PCs.
Thus, a \emph{distributed memory} parallel system
is generally required for performing DNS.
Here, the traditional approach has been
to distribute parallel tasks among individual CPU cores,
or groups of CPU cores
that share common memory spaces.
In the latter case,
each group of cores constitutes a \emph{shared memory} system,
and the overall system is referred to as a \emph{hybrid system}.

@@

      f^{\prime}_2 \\
      f^{\prime}_3 \\
      \vdots \\
      \vdots \\
      \vdots \\
      \vdots \\
      f^{\prime}_{n-1} \\
      f^{\prime}_n
   \end{bmatrix}
 =
 \begin{bmatrix}
     \frac{-5f_1 + 4f_2 + f_3}{2dx}\\
     \frac{3(f_{3} - f_{1})}{4dx}\\
     \frac{3(f_{4} - f_{2})}{4dx}\\
     \vdots\\
     \vdots\\
     \vdots\\
     \vdots\\
     \frac{3(f_{n} - f_{n-2})}{4dx}\\
     \frac{5f_{n} - 4f_{n-1} - f_{n-2}}{2dx}
  \end{bmatrix}
\end{equation}

Similar tridiagonal schemes are available
to evaluate higher order derivatives.
Compact schemes for
second, third and fourth derivatives
are derived in \cite{lele1992compact}.

@@

As described,
the CUDA programming interface
allows developing general-purpose GPU applications---however,
there are other options.
Notably,
the OpenACC \cite{openacc} toolkit allows
portions of code to be offloaded to GPU accelerators
simply by use of compiler directives.
This makes porting existing applications
extremely easy,
and introduces almost no extra code.
Some libraries for scientific computation,
such as PETSc \cite{petsc-web-page},
also allow the \emph{transparent} use of GPUs
through their interfaces.
We use the CUDA interface because
it fully exposes the
features of the underlying architecture
that can be exploited.
This is especially important due to the
complex computational and memory access patterns involved
in our algorithms.

>>>>

When solving for flows numerically,
at each of the grid points in the computational domain,
various data about the flow must be stored---for
example, the geometric coordinates ($x$, $y$ and $z$),
pressure, temperature and velocities at that point.
For even small problem sizes,
the amount of memory required to store this data
can quickly exceed the capacity of modern workstations/PCs.
Thus, a \emph{distributed memory} parallel system
is generally required for performing DNS.
Here, the traditional approach has been
to distribute parallel tasks among individual CPU cores,
or groups of CPU cores
that share common memory spaces.
In the latter case,
each group of cores constitutes a \emph{shared memory} system,
and the overall system is referred to as a \emph{hybrid system}.

@@

      f^{\prime}_2 \\
      f^{\prime}_3 \\
      \vdots \\
      \vdots \\
      \vdots \\
      \vdots \\
      f^{\prime}_{n-1} \\
      f^{\prime}_n
   \end{bmatrix}
 =
 \begin{bmatrix}
     \frac{-5f_1 + 4f_2 + f_3}{2dx}\\
     \frac{3(f_{3} - f_{1})}{4dx}\\
     \frac{3(f_{4} - f_{2})}{4dx}\\
     \vdots\\
     \vdots\\
     \vdots\\
     \vdots\\
     \frac{3(f_{n} - f_{n-2})}{4dx}\\
     \frac{5f_{n} - 4f_{n-1} - f_{n-2}}{2dx}
  \end{bmatrix}
\end{equation}
%
Similar tridiagonal schemes are available
to evaluate higher order derivatives.
Compact schemes for
second, third and fourth derivatives
are derived in \cite{lele1992compact}.

@@

As described,
the CUDA programming interface
allows developing general-purpose GPU applications---however,
there are other options.
The OpenCL framework \cite{opencl}
is used to write applications
that can be ported across
a variety architectures including NVIDIA GPUs.
and introduces almost no extra code.
Several GPU accelerated libraries \cite{gpulibraries}
allow existing applications to be ported to GPUs
by offering compatibility with industry standard CPU libraries.
The OpenACC \cite{openacc} toolkit allows
portions of code to be offloaded to GPU accelerators
simply by use of compiler directives,
similar to OpenMP.
This makes porting existing applications
extremely easy.
Many frameworks for scientific computing
such at PETSc \cite{petsc-web-page}
also allow the use of GPUs
for most of their functionality.

We use the CUDA interface because
it fully exposes the
features of the underlying architecture
that can be exploited.
This is especially important due to the
complex computational and memory access patterns involved
in our algorithms.

### 2015-11-24 thesis/proposed-tridiagonal-algorithm.tex
some more updates to thesis

In this section, we describe the implementation of a GPU solver
for solving a given \emph{near-Toeplitz} tridiagonal system
for multiple right hand sides.
Our implementation uses the
NVIDIA CUDA platform for programming the GPU,
but is easily translated to OpenCL.
The Python programming language is used
to interface with CUDA, by use of the
PyCUDA \cite{kloeckner_pycuda_2012} library.
We develop two approaches based on the common idea
of precomputed forward reduction coefficients---one that leverages
the GPU's \emph{shared memory},
and the other working directly with global memory.
The GPU kernels for both implementations are relatively
straightforward and compact,
spanning no more than 100 lines of code.

@@

Figure \ref{fig:cyclic-reduction-precomputing}
shows the
\emph{maximum} amount of storage required for
storing each of the coefficients
$\{b_1, c_1, a_0, b_0, c_0, a_n, b_n\}$,
or auxiliary variables $k_1$ and $k_2$.
A careful analysis of the
forward reduction and backward substitution equations
reveals that the actual amount of storage is somewhat less.
For instance,
the values $b_n^m$ are unused
in Eqs (\ref{eqn:precomputed-forward-reduction-step})-
(\ref{eqn:precomputed-backward-substitution-step-1}).
The values $a_n^m$ and $b_n^m$ are
required only for solving the $2-by-2$
system at the end of the forward reduction phase,
i.e., for $m=log_2(n)-1$,
and they are not stored for the previous steps.
Similarly, the values $c_1^m$ are equal to $c_0^m$,
and do not require separate storage.
The set of precomputed coefficients required to be stored is then
$\{a_0^m, b_0^m, c_0^m, k_1^m, k_2^m, b_1^m, k_{1,1}^m, k_{1,n}^m\}$.
Each of these ``coefficient arrays'' are computed on the CPU
and transferred to the GPU.
Additionally, the two scalars $a_n^{log_2(n)-1}$ and $b_n^{log_2(n)-1}$
are required for the 2-by-2 solve.
%
\begin{figure}
\begin{center}
\includegraphics[height=200pt]{img/block-mapping.eps}
\end{center}
\caption{Storing right hand sides and mapping to thread blocks.}
\label{fig:block-mapping}
\end{figure}
%
The right hand sides that the system must be solved for
are stored in a single contiguous array in GPU memory
\ref{fig:block-mapping}.
The precomputed coefficient arrays and the right hand side array
are passed as inputs to the compute kernels that implement
the modified cyclic reduction.
We describe two implementations of the kernels,
the first works entirely with global memory,
and the second leverages the GPU's \emph{shared} memory.

>>>>

In this section, we describe the implementation of a GPU solver
for solving a given \emph{near-Toeplitz} tridiagonal system
for multiple right hand sides.
Our implementation uses the
NVIDIA CUDA platform for programming the GPU,
but is easily translated to OpenCL.
The Python programming language is used
to interface with CUDA, by use of the
PyCUDA \cite{kloeckner_pycuda_2012} library.
We develop two approaches based on the common idea
of precomputed forward reduction coefficients---one that leverages
the GPU's \emph{shared memory},
and the other working directly with global memory.
The GPU kernels for both implementations are relatively
straightforward and compact,
spanning no more than 100 lines of code
(see Appendix).

@@

Figure \ref{fig:cyclic-reduction-precomputing}
shows the
\emph{maximum} amount of storage required for
storing each of the coefficients
$\{b_1, c_1, a_0, b_0, c_0, a_n, b_n\}$,
or auxiliary variables $k_1$ and $k_2$.
A careful analysis of the
forward reduction and backward substitution equations
reveals that the actual amount of storage is somewhat less.
For instance,
the values $b_n^m$ are unused
in Eqs (\ref{eqn:precomputed-forward-reduction-step}) -
(\ref{eqn:precomputed-backward-substitution-step-1}).
The values $a_n^m$ and $b_n^m$ are
required only for solving the $2-by-2$
system at the end of the forward reduction phase,
i.e., for $m=log_2(n)-1$,
and they are not stored for the previous steps.
Similarly, the values $c_1^m$ are equal to $c_0^m$,
and do not require separate storage.
The set of precomputed coefficients required to be stored is then
$\{a_0^m, b_0^m, c_0^m, k_1^m, k_2^m, b_1^m, k_{1,1}^m, k_{1,n}^m\}$.
Each of these ``coefficient arrays'' are computed on the CPU
and transferred to the GPU.
Additionally, the two scalars $a_n^{log_2(n)-1}$ and $b_n^{log_2(n)-1}$
are required for the 2-by-2 solve.
%
\begin{figure}
\begin{center}
\includegraphics[height=200pt]{img/block-mapping.eps}
\end{center}
\caption{Storing right hand sides and mapping to thread blocks.}
\label{fig:block-mapping}
\end{figure}
%
The right hand sides that the system must be solved for
are stored in a single contiguous array in GPU memory
\ref{fig:block-mapping}.
The precomputed coefficient arrays and the right hand side array
are passed as inputs to the compute kernels that implement
the modified cyclic reduction.
We describe two implementations of the kernels,
the first works entirely with global memory,
and the second leverages the GPU's \emph{shared} memory.

### 2015-11-24 thesis/results.tex
some more updates to thesis

In Fig. \ref{fig:bench-2d} and \ref{fig:bench-3d},
we provide the relative performance of
Intel MKL and CUSPARSE solvers and compare against
the NEATO shared memory implementation.
The relative performance for
each problem size is obtained by
normalizing the solver timings
by the timing for the NEATO solver for that problem size.
Table \ref{table:bench} shows the timings of the various
solvers to solve different problem sizes.

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=300pt]{fig/profiling-1024-64.eps}
\caption{Solving problem sized $1024^3$ on 64 GPUs}
\label{fig:compact-profiling-1024-64}
\end{center}
\end{figure}
%
\begin{figure}[h!]
\begin{center}
\includegraphics[width=300pt]{fig/profiling-2048-64.eps}
\caption{Solving problem sized $2048^3$ on 64 GPUs}
\label{fig:compact-profiling-2048-64}
\end{center}
\end{figure}
%
Figures \ref{fig:compact-profiling-1024-64}-\ref{fig:compact-profiling-2048-64}
show the time taken by the different steps of
the compact finite difference solver.
We note that for the larger problem size,
the evaluation of the primary systems (using the NEATO solver)
constitutes a larger majority of the total runtime,
which justifies our efforts in optimizing the tridiagonal solver.
For evaluation of the derivatives in the y- and z- directions,
we note that a significant portion of the runtime is dedicated
to performing permutations of the input and output data.
We attribute this to our na\"{\i}ve implementation of the
permutation kernels (no shared memory).
%
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/strong-scaling-256.eps}
\caption{Strong scaling for multi-GPU compact finite difference, problem size: $256^3$.}
\label{fig:strong-scaling-256}
\end{center}
\end{figure}
%
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/strong-scaling-512.eps}
\caption{Strong scaling for multi-GPU compact finite difference, problem size: $512^3$.}
\label{fig:strong-scaling-512}
\end{center}
\end{figure}
%
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/weak-scaling-128.eps}
\caption{Weak scaling for multi-GPU compact finite difference, problem size: $128^3$ per process.}
\label{fig:weak-scaling-128}
\end{center}
\end{figure}
%
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/weak-scaling-256.eps}
\caption{Weak scaling for multi-GPU compact finite difference, problem size: $256^3$ per process.}
\label{fig:weak-scaling-256}
\end{center}
\end{figure}
%
Figures \ref{fig:strong-scaling-256}-\ref{fig:weak-scaling-256}
show the strong and weak scaling of the compact finite difference solver
for evaluating derivatives in all three coordinate directions.
We note good scaling for both 2-D and 3-D cases.
The strong scaling for 3-D problems is somewhat better
as the GPU is kept busy for all problem sizes.
%
\input{table/compact-refimpl-timings.tex}
%
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/compact-refimpl-speedups.eps}
\caption{Speedups over reference implementation}
\label{fig:compact-refimpl-speedups}
\end{center}
\end{figure}
%
We also compare the performance of our compact finite difference
solver with the implementation by
Mohd-Yusof et al. \cite{mohd2010adapting}.
This implementation uses a
distributed tridiagonal solver based on the LU algorithm
to solve the linear systems.
For comparing timings,
we use the number of CPU sockets as the basis.
Each CPU socket uses 8 CPU cores and 1 GPU.
Thus we maintain a ratio of 1:8 between GPUs and CPU cores
in our comparison.
Table \ref{table:compact-refimpl-timings}
shows timings for computing derivatives for problem
sized up to $2048^3$,
and Fig. \ref{fig:compact-refimpl-speedups}
shows the respective speedup using our implementation.

>>>>

In Fig. \ref{fig:bench-2d} and \ref{fig:bench-3d},
we provide the relative performance of
Intel MKL and CUSPARSE solvers and compare against
the NEATO shared memory implementation.
The relative performance for
each problem size is obtained by
normalizing the solver timings
by the timing for the NEATO solver for that problem size.
Table \ref{table:bench} shows the timings of the various
solvers to solve different problem sizes.
Note that data is missing for the CUSPARSE solver
for the $512^3$ 3-D case,
as the GPU was unable to accomodate this problem size---this
is due to the large amount of scratch space required
by the CUSPARSE implementation.

@@

\subsection{Performance profiling}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=300pt]{fig/profiling-1024-64.eps}
\caption{Solving problem sized $1024^3$ on 64 GPUs}
\label{fig:compact-profiling-1024-64}
\end{center}
\end{figure}
%
\begin{figure}[h!]
\begin{center}
\includegraphics[width=300pt]{fig/profiling-2048-64.eps}
\caption{Solving problem sized $2048^3$ on 64 GPUs}
\label{fig:compact-profiling-2048-64}
\end{center}
\end{figure}
%
Figures \ref{fig:compact-profiling-1024-64} - \ref{fig:compact-profiling-2048-64}
show the time taken by the different steps of
the compact finite difference solver.
We note that for the larger problem size,
the evaluation of the primary systems (using the NEATO solver)
constitutes a larger majority of the total runtime,
which justifies our efforts in optimizing the tridiagonal solver.
For evaluation of the derivatives in the $y-$ and $z-$ directions,
we note that a significant portion of the runtime is dedicated
to performing permutations of the input and output data.
We attribute this to our na\"{\i}ve implementation of the
permutation kernels (no shared memory usage).

\subsection{Strong and weak scaling}
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/strong-scaling-256.eps}
\caption{Strong scaling for multi-GPU compact finite difference, problem size: $256^3$.}
\label{fig:strong-scaling-256}
\end{center}
\end{figure}
%
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/strong-scaling-512.eps}
\caption{Strong scaling for multi-GPU compact finite difference, problem size: $512^3$.}
\label{fig:strong-scaling-512}
\end{center}
\end{figure}
%
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/weak-scaling-128.eps}
\caption{Weak scaling for multi-GPU compact finite difference, problem size: $128^3$ per process.}
\label{fig:weak-scaling-128}
\end{center}
\end{figure}
%
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/weak-scaling-256.eps}
\caption{Weak scaling for multi-GPU compact finite difference, problem size: $256^3$ per process.}
\label{fig:weak-scaling-256}
\end{center}
\end{figure}
%
Figures \ref{fig:strong-scaling-256} - \ref{fig:weak-scaling-256}
show the strong and weak scaling of the compact finite difference solver
for evaluating derivatives in all three coordinate directions.
For the strong scaling measurement,
we keep the problem size fixed
and increase the number of GPUs used to solve the problem.
For the weak scaling measurement,
we keep the problem size \emph{per GPU}
fixed, and increase the number of GPUs used.
We note good scaling for both 2-D and 3-D cases.
The strong scaling for 3-D problems is somewhat better
as the GPU is kept busy for all problem sizes.

\subsection{Comparison with a CPU-only approach}

\input{table/compact-refimpl-timings.tex}
%
\begin{figure}
\begin{center}
\includegraphics[width=300pt]{fig/compact-refimpl-speedups.eps}
\caption{Speedups over reference implementation
    for computing derivative in the fastest coordinate direction}
\label{fig:compact-refimpl-speedups}
\end{center}
\end{figure}

We also compare the performance of our compact finite difference
solver with the approach described by
Mohd-Yusof et al.\ \cite{mohd2010adapting},
implemented for CPUs.
The approach uses a distributed tridiagonal solver based on
the LU decomposition specialized for tridiagonal systems.
The problem is divided among individual CPU cores,
communicating via MPI.
For comparing timings,
we use the number of CPU sockets as the basis.
Each CPU socket uses 8 CPU cores and 1 GPU.
Thus, we maintain a ratio of 1:8 between GPUs and CPU cores
in our comparison.
Table \ref{table:compact-refimpl-timings}
shows timings for computing derivatives
in the fastest coordinate direction
for problems sized up to $2048^3$,
and Fig. \ref{fig:compact-refimpl-speedups}
shows the respective speedup using our implementation.
